Abstraction Refinement for Emptiness Checking of Alternating Data Automata

### 2018-02-22 tacas/draft.tex
final

\begin{abstract}
Alternating automata have been widely used to model and verify systems
that handle data from finite domains, such as communication protocols
or hardware. The main advantage of the alternating model of
computation is that complementation is possible in linear time, thus
allowing to concisely encode trace inclusion problems that occur often
in verification. In this paper we consider alternating automata over
infinite alphabets, whose transition rules are formulae in a combined
theory of booleans and some infinite data domain, that relate past and
current values of the data variables. The data theory is not fixed,
but rather it is a parameter of the class. We show that union,
intersection and complementation are possible in linear time in this
model and, though the emptiness problem is undecidable, we provide two
efficient semi-algorithms, inspired by two state-of-the-art
abstraction refinement model checking methods: lazy predicate
abstraction \cite{HJMS02} and the \impact~ semi-algorithm
\cite{mcmillan06}. We have implemented both methods and report the
results of an experimental comparison.
\end{abstract}

@@

In this paper, we present a new model of alternating automata over
infinite alphabets consisting of pairs $(a,\nu)$ where $a$ is an input
event from a finite set and $\nu$ is a valuation of a finite set
$\vec{x}$ of variables that range over an infinite domain. We assume
that, at all times, the successive values taken by the variables in
$\vec{x}$ are an observable part of the language, in other words,
there are no hidden variables in our model. The transition rules are
specified by a set of formulae, in a combined first-order theory of
boolean control states and data, that relate past with present values
of the variables. We do not fix the data theory a~priori, but rather
consider it to be a parameter of the class.

A run over an input word $(a_1,\nu_1) \ldots (a_n,\nu_n)$ is a
sequence $\phi_0(\vec{x}_0) \Rightarrow \phi_1(\vec{x}_0,\vec{x}_1)
\Rightarrow \ldots \Rightarrow \phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ of
rewritings of the initial formula by substituting boolean states with
time-stamped transition rules. The word is accepted if the final
formula $\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ holds, when all
time-stamped variables $\vec{x}_1,\ldots,\vec{x}_n$ are substituted by
their values in $\nu_1,\ldots,\nu_n$, all non-final states replaced by
false and all final states by true.

The boolean operations of union, intersection and complement can be
implemented in linear time in this model, thus matching the complexity
of performing these operations in the finite-alphabet case. The price
to be paid is that emptiness becomes undecidable, for which reason we
provide two efficient semi-algorithms for emptiness, based on lazy
predicate abstraction \cite{HJMS02} and the \impact~method
\cite{mcmillan06}.  These algorithms are proven to terminate and
return a word from the language of the automaton, if one exists, but
termination is not guaranteed when the language is empty.

%% Finally, we prove termination in the empty case, when the transition
%% formulae\begin{inparaenum}[(i)]
%% \item contain only constants of the data sort and boolean functions of
%%   any arity, or they
%% %
%% \item constrain only the current but not the past values on the input,
%%   as in the case of symbolic automata \cite{symbTransd:POPL12,DAntoniKW16}.
%% \end{inparaenum}

We have implemented the boolean operations and emptiness checking
semi-algorithms and carried out experiments with examples taken from
array logics \cite{cav09}, timed automata
\cite{henzinger:RealTimeSystems}, communication protocols \cite{abp}
and hardware verification \cite{smrcka}.

@@

The model presented in this paper stems from the language inclusion
problem considered in \cite{Tacas16}. There we provide a
semi-algorithm for inclusion of data languages, based on an
exponential determinization procedure and an abstraction refinement
loop using lazy predicate abstraction \cite{HJMS02}. In this work we
consider the full model of alternation and rely entirely on the
ability of SMT solvers to produce interpolants in the combined theory
of booleans and data. Since determinisation is not needed and
complementation is possible in linear time, the bulk of the work is
carried out by the solver.

@@

Given sets of variables $\vec{b},\vec{x} \subset \vars$ of sort
$\Bool$ and $\Data$, respectively, we denote by
$\Form(\vec{b},\vec{x})$ the set of formulae $\phi$ such that
$\fv{\Bool}{\phi} \subseteq \vec{b}$ and $\fv{\Data}{\phi} \subseteq
\vec{x}$. By $\Form^+(\vec{b},\vec{x})$ we denote the set of formulae
from $\Form(\vec{b},\vec{x})$ in which each boolean variable occurs
under an even number of negations.

@@

\begin{figure}[htb]
\begin{center}
\begin{minipage}{6cm}
\centerline{\includegraphics[scale=0.8]{Example.pdf}}
\centerline{(a)}
\end{minipage}
\begin{minipage}{6cm}
\[\begin{array}{rcl}
\Delta(q_0,a) & \equiv & q_1 \wedge q_2 \wedge x\teq0 \wedge y\teq0 \\
\Delta(q_1,a) & \equiv & q_1 \wedge q_3 \wedge x\teq\overline{y}+1 \wedge y\teq\overline{x}+1 \\
\Delta(q_1,b) & \equiv & q_3 \wedge \overline{x} \geq \overline{y} \\
\Delta(q_2,a) & \equiv & q_2 \wedge x > \overline{x} \wedge y > \overline{y} \\
\Delta(q_2,b) & \equiv & q_4 \wedge \overline{x}>\overline{y} \\\\
\end{array}\]
\centerline{(b)}
\end{minipage}
\end{center}
\caption{Alternating Data Automaton Example}
\label{fig:ex}
\end{figure}

@@

In this paper we tackle the following problems: \begin{compactenum}
\item \emph{boolean closure}: given automata $\A_1$ and $\A_2$, both
  with the same set of variables $\vec{x}$, do there exist automata
  $\A_\cup$, $\A_\cap$ and $\overline{\A_1}$ such that $L(\A_\cup) =
  \A_1 \cup \A_2$, $L(A_\cap) = \A_1 \cap \A_2$ and
  $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$ ?
\end{compactenum}

@@

Given a set $Q$ of boolean variables and a set $\vec{x}$ of variables
of sort $\Data$, for a formula $\phi \in \Form^+(Q,\vec{x})$, with no
negated occurrences of the boolean variables, we define the formula
$\overline{\phi} \in \Form^+(Q,\vec{x})$ recursively on the structure
of $\phi$:
\[\begin{array}{rclcrcl}
\overline{\phi_1 \vee \phi_2} & \equiv & \overline{\phi_1} \wedge \overline{\phi_2} && 
\overline{\phi_1 \wedge \phi_2} & \equiv & \overline{\phi_1} \vee \overline{\phi_2} \\ 
\overline{\neg\phi} & \equiv & \neg \overline{\phi} \text{ if $\phi$ not atom} &&
\overline{\phi} & \equiv & \phi \text{ if $\phi \in Q$} \\
\overline{\phi} & \equiv & \neg\phi \text{ if $\phi \not\in Q$ atom}
\end{array}\]
We have $\len{\overline{\phi}} = \len{\phi}$, for every formula $\phi
\in \Form^+(Q,\vec{x})$.

@@

It is easy to see that $\len{\A_\cup} = \len{\A_\cap} = \len{\A_1} +
\len{\A_2}$ and $\len{\overline{\A}} = \len{\A}$, thus the automata
for the boolean operations, including complementation, can be built in
linear time. This matches the linear-time bounds for intersection and
complementation of alternating automata over finite alphabets
\cite{ChandraKozenStockmeyer81}.

@@

Since no boolean variable from $Q$ occurs under negation in $\phi$, it
is easy to prove the following monotonicity property: given two
formulae $\phi,\psi \in \Form^+(Q,\vec{x})$ if $\phi \models \psi$
then $\Post{\A}(\phi,u) \models \Post{\A}(\psi,u)$, for any $u \in
\Sigma^*$. This suggest an improvement of the above semi-algorithm,
that enumerates and stores only a set $U \subseteq \Sigma^*$ for which
$\set{\Post{\A}(\phi,u) \mid u \in U}$ forms an
\emph{antichain}\footnote{Given a partial order $(D,\preceq)$ an
  antichain is a set $A \subseteq D$ such that $a \not\preceq b$ for
  any $a,b \in A$.} w.r.t. the entailment partial order. This is
because, for any $u,v \in \Sigma^*$, if $\Post{\A}(\iota,u) \models
\Post{\A}(\iota,v)$ and $\Accept{\A}(uw)$ is satisfiable for some $w
\in \Sigma^*$, then $\Post{\A}(\iota,uw) \models \Post{\A}(\iota,vw)$,
thus $\Accept{\A}(vw)$ is satisfiable as well, and there is no need
for $u$, since the non-emptiness of $\A$ can be proved using $v$
alone. However, even with this optimization, the enumeration of
sequences from $\Sigma^*$ diverges in many real cases, because
infinite antichains exist in many interpretations, e.g.\ $q \wedge x
\teq 0,~ q \wedge x \teq 1, \ldots$ for $\Data^\I = \nat$.

A \emph{safety invariant} for $\A$ is a function $\Inv : (Q \mapsto
\booli) \rightarrow 2^{\vec{x} \mapsto \Data^\I}$ such that, for every
boolean valuation $\beta : Q \rightarrow \booli$, every valuation $\nu
: \vec{x} \mapsto \Data^\I$ of the data variables and every finite
sequence $u \in \Sigma^*$ of input events, the following
hold: \begin{compactenum}
%% \item $\beta \models \iota \Rightarrow \nu_0 \in \Inv(\beta)$, where
%%   $\nu_0(x) = \mathbf{0}$, for all $x \in \vec{x}$.
%
\item\label{it1:inv} $\I,\beta \cup \nu \models \Post{\A}(\iota,u)
  \Rightarrow \nu \in \Inv(\beta)$, and
%
\item\label{it2:inv} $\nu \in \Inv(\beta) \Rightarrow \I,\beta \cup
  \nu \not\models \Accept{\A}(u)$.
\end{compactenum}
If $\Inv$ satisfies only the first point above, we call it an
\emph{invariant}. Intuitively, a safety invariant maps every boolean
valuation into a set of data valuations, that contains the initial
configuration $\iota \equiv \Post{\A}(\iota,\varepsilon)$, whose data
variables are unconstrained, over-approximates the set of reachable
valuations (point \ref{it1:inv}) and excludes the valuations satisfying the
acceptance condition (point \ref{it2:inv}).  A formula $\phi(Q,\vec{x})$
is said to \emph{define} $\Inv$ iff for all $\beta: Q \rightarrow \booli$ and
$\nu : \vec{x} \rightarrow \Data^\I$, we have $\I,\beta\cup\nu \models
\phi$ iff $\nu \in \Inv(\beta)$.

@@

We are left with fulfilling point (\ref{it2:inv}) from the definition
of a safety invariant. To this end, suppose that, for a given set
$\preds$ of predicates, the invariant $\abs{\Inv}$, defined by the
previous lemma, meets point (\ref{it1:inv}) but not point
(\ref{it2:inv}), where $\Post{\A}$ and $\Accept{\A}$ replace
$\AbsPost{\A}$ and $\AbsAccept{\A}$, respectively. In other words,
there exists a finite sequence $u \in \Sigma^*$ such that $\nu \in
\abs{\Inv}(\beta)$ and $\I,\beta\cup\nu \models \AbsAccept{\A}(u)$,
for some boolean $\beta : Q \rightarrow \booli$ and data $\nu :
\vec{x} \rightarrow \Data^\I$ valuations. Such a $u\in\Sigma^*$ is
called a \emph{counterexample}.

@@

Given an unsatisfiable conjunction of formulae $\psi_1 \wedge \ldots
\wedge \psi_n$, an \emph{interpolant} is a tuple of formulae
$\tuple{I_1,\ldots,I_{n-1},I_n}$ such that $I_n \equiv \bot$, $I_i
\wedge \psi_i \models_\T I_{i+1}$ and $I_i$ contains only variables
and function symbols that are common to $\psi_i$ and $\psi_{i+1}$, for
all $i \in [n-1]$. Moreover, by Lyndon's Interpolation Theorem
\cite{Lyndon59}, we can assume without loss of generality that every
boolean variable with at least one positive (negative) occurrence in
$I_i$ has at least one positive (negative) occurrence in both $\psi_i$
and $\psi_{i+1}$. In the following, we shall assume the existence of
an interpolating decision procedure for $\theory(\sig,\I)$ that meets
the requirements of Lyndon's Interpolation Theorem.

@@

Given a non-empty spurious counterexample $u = a_1\ldots a_n$, where
$n>0$, we consider the following interpolation problem: 
\begin{eqnarray}\label{eq:interpolation-problem}
\Theta(u) & \equiv & \theta_0(Q_0) \wedge \theta_1(Q_0 \cup Q_1,\vec{x}_0
\cup \vec{x}_1) \wedge \ldots \\ 
&& \wedge~ \theta_n(Q_{n-1} \cup Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge \theta_{n+1}(Q_n) \nonumber
\end{eqnarray}
where $Q_k = \set{q_k \mid q \in Q}$, $k \in [0,n]$ are time-stamped
sets of boolean variables corresponding to the set $Q$ of states of
$\A$. The first conjunct $\theta_0(Q_0) \equiv \iota[Q_0/Q]$ is the
initial configuration of $\A$, with every $q \in \fv{\Bool}{\iota}$
replaced by $q_0$. The definition of $\theta_k$, for all $k\in[1,n]$,
uses \emph{replacement sets} $R_\ell \subseteq Q_\ell$, $\ell\in
[0,n]$, which are defined inductively below: \begin{compactitem}
\item $R_0 = \fv{\Bool}{\theta_0}$,
%
\item $\theta_\ell \equiv \bigwedge_{q_{\ell-1}\in R_{\ell-1}}
  (q_{\ell-1} \rightarrow
  \Delta(q,a_\ell)[Q_\ell/Q,\vec{x}_{\ell-1}/\overline{\vec{x}},\vec{x}_\ell/\vec{x}])$
  and $R_\ell = \fv{\Bool}{\theta_\ell} \cap Q_\ell$, for each
  $\ell\in[1,n]$.
%
\item $\theta_{n+1}(Q_n) \equiv \bigwedge_{q \in Q \setminus F} (q_n
  \rightarrow \bot)$.
\end{compactitem}
The intuition is that $R_0,\ldots,R_n$ are the sets of states
replaced, $\theta_0, \ldots, \theta_n$ are the sets of transition
rules fired on the run of $\A$ over $u$ and $\theta_{n+1}$ is the
acceptance condition, which forces the last remaining non-final states
to be false.
%
%% Moreover, we require that an interpolant
%% $\tuple{\top,I_0,\ldots,I_n,\bot}$ for the interpolation problem
%% $\Theta(u)$ does not have negative occurrences of states, i.e.\ $I_i
%% \in \Form^+(Q,\vec{x})$, for all $i \in [0,n]$. Such an interpolant
%% can always be built, as showed below:
%
%% \begin{proposition}\label{prop:positive-interpol}
%%   If $\Theta(a_1 \ldots a_n)$ is unsatisfiable then one can build an
%%   interpolant $\tuple{\top,I_0,\ldots,I_n,\bot}$ for $\Theta(a_1\ldots
%%   a_n)$, such that $I_i \in \Form^+(Q,\vec{x})$, for all $i \in
%%   [0,n]$.
%% \end{proposition}
%
We recall that a run of $\A$ over $u$ is a sequence: \[\phi_0(Q)
\Rightarrow \phi_1(Q,\vec{x}_0\cup\vec{x}_1) \Rightarrow \ldots
\Rightarrow \phi_n(Q,\vec{x}_0\cup\ldots\cup\vec{x}_n)\] where
$\phi_0$ is the initial configuration $\iota$ and for each $k>0$,
$\phi_k$ is obtained from $\phi_{k-1}$ by replacing each state $q \in
\fv{\Bool}{\phi_{k-1}}$ by the formula
$\Delta(q,a_k)[\vec{x}_{k-1}/\overline{\vec{x}},\vec{x}_k/\vec{x}]$,
given by the transition function of $\A$. Observe that, because the
states are replaced with transition formulae when moving one step in a
run, these formulae lose track of the control history and are not
suitable for producing interpolants that relate states and data.

The main idea behind the above definition of the interpolation problem
is that we would like to obtain an interpolant $\tuple{\top,I_0(Q),
  I_1(Q,\vec{x}), \ldots, I_{n}(Q,\vec{x}),\bot}$ whose formulae
\emph{combine states with the data constraints that must hold
  locally}, whenever the control reaches a certain boolean
configuration. This association of states with data valuations is
tantamount to defining efficient semi-algorithms, based on lazy
abstraction \cite{HJMS02}. Furthermore, the abstraction defined by the
interpolants generated in this way can also \emph{over-approximate the
  control structure} of an automaton, in addition to the sets of data
values encountered throughout its runs.

@@

Termination of Algorithm \ref{alg:predabs} depends on the ability of a
given interpolating decision procedure for the combined boolean and
data theory $\theory(\sig,\I)$ to provide interpolants that yield a
safety invariant, whenever $L(\A) = \emptyset$. In this case, we use
the covering relation $\lhd$ to ensure that, when a newly generated
node is covered by a node already in $N$, it is not added to the
worklist, thus cutting the current branch of the search.

@@

In this process, the nodes on the path between $\rootNode$ and $n$,
including $n$, might become eligible for coverage, therefore we
attempt to close each ancestor of $n$ that is impacted by the
refinement (line \ref{ln:refine-end}). Observe that, in this case the
call to $\Call{Close}{}$ must uncover each node which is covered by a
successor of $n$ (line \ref{ln:close-uncover} of the $\Call{Close}{}$
function). This is required because, due to the over-approximation of
the sets of reachable configurations, the covering relation is not
transitive, as explained in \cite{mcmillan06}. If $\Call{Close}{}$
adds a covering edge $(n_i,m)$ to $\lhd$, it does not have to be
called for the successors of $n_i$ on this path, which is handled via
the boolean flag $b$.

>>>>

\begin{abstract}
Alternating automata have been widely used to model and verify systems
that handle data from finite domains, such as communication protocols
or hardware. The main advantage of the alternating model of
computation is that complementation is possible in linear time, thus
allowing to concisely encode trace inclusion problems that occur often
in verification. In this paper we consider alternating automata over
infinite alphabets, whose transition rules are formulae in a combined
theory of Booleans and some infinite data domain, that relate past and
current values of the data variables. The data theory is not fixed,
but rather it is a parameter of the class. We show that union,
intersection and complementation are possible in linear time in this
model and, though the emptiness problem is undecidable, we provide two
efficient semi-algorithms, inspired by two state-of-the-art
abstraction refinement model checking methods: lazy predicate
abstraction \cite{HJMS02} and the \impact~ semi-algorithm
\cite{mcmillan06}. We have implemented both methods and report the
results of an experimental comparison.
\end{abstract}

@@

In this paper, we present a new model of alternating automata over
infinite alphabets consisting of pairs $(a,\nu)$ where $a$ is an input
event from a finite set and $\nu$ is a valuation of a finite set
$\vec{x}$ of variables that range over an infinite domain. We assume
that, at all times, the successive values taken by the variables in
$\vec{x}$ are an observable part of the language, in other words,
there are no hidden variables in our model. The transition rules are
specified by a set of formulae, in a combined first-order theory of
Boolean control states and data, that relate past with present values
of the variables. We do not fix the data theory a~priori, but rather
consider it to be a parameter of the class.

A run over an input word $(a_1,\nu_1) \ldots (a_n,\nu_n)$ is a
sequence $\phi_0(\vec{x}_0) \Rightarrow \phi_1(\vec{x}_0,\vec{x}_1)
\Rightarrow \ldots \Rightarrow \phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ of
rewritings of the initial formula by substituting Boolean states with
time-stamped transition rules. The word is accepted if the final
formula $\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ holds, when all
time-stamped variables $\vec{x}_1,\ldots,\vec{x}_n$ are substituted by
their values in $\nu_1,\ldots,\nu_n$, all non-final states replaced by
false and all final states by true.

The Boolean operations of union, intersection and complement can be
implemented in linear time in this model, thus matching the complexity
of performing these operations in the finite-alphabet case. The price
to be paid is that emptiness becomes undecidable, for which reason we
provide two efficient semi-algorithms for emptiness, based on lazy
predicate abstraction \cite{HJMS02} and the \impact~method
\cite{mcmillan06}.  These algorithms are proven to terminate and
return a word from the language of the automaton, if one exists, but
termination is not guaranteed when the language is empty.

%% Finally, we prove termination in the empty case, when the transition
%% formulae\begin{inparaenum}[(i)]
%% \item contain only constants of the data sort and Boolean functions of
%%   any arity, or they
%% %
%% \item constrain only the current but not the past values on the input,
%%   as in the case of symbolic automata \cite{symbTransd:POPL12,DAntoniKW16}.
%% \end{inparaenum}

We have implemented the Boolean operations and emptiness checking
semi-algorithms and carried out experiments with examples taken from
array logics \cite{cav09}, timed automata
\cite{henzinger:RealTimeSystems}, communication protocols \cite{abp}
and hardware verification \cite{smrcka}.

@@

The model presented in this paper stems from the language inclusion
problem considered in \cite{Tacas16}. There we provide a
semi-algorithm for inclusion of data languages, based on an
exponential determinization procedure and an abstraction refinement
loop using lazy predicate abstraction \cite{HJMS02}. In this work we
consider the full model of alternation and rely entirely on the
ability of SMT solvers to produce interpolants in the combined theory
of Booleans and data. Since determinisation is not needed and
complementation is possible in linear time, the bulk of the work is
carried out by the solver.

@@

Given sets of variables $\vec{b},\vec{x} \subset \vars$ of sort
$\Bool$ and $\Data$, respectively, we denote by
$\Form(\vec{b},\vec{x})$ the set of formulae $\phi$ such that
$\fv{\Bool}{\phi} \subseteq \vec{b}$ and $\fv{\Data}{\phi} \subseteq
\vec{x}$. By $\Form^+(\vec{b},\vec{x})$ we denote the set of formulae
from $\Form(\vec{b},\vec{x})$ in which each Boolean variable occurs
under an even number of negations.

@@

\begin{figure}[htb]
\begin{center}
\begin{minipage}{6cm}
\centerline{\includegraphics[scale=0.8]{Example.pdf}}
\centerline{(a)}
\end{minipage}
\begin{minipage}{6cm}
\[\begin{array}{rcl}
\Delta(q_0,a) & \equiv & q_1 \wedge q_2 \wedge x\teq0 \wedge y\teq0 \\
\Delta(q_1,a) & \equiv & q_1 \wedge q_2 \wedge x\teq\overline{y}+1 \wedge y\teq\overline{x}+1 \\
\Delta(q_1,b) & \equiv & q_3 \wedge \overline{x} \geq \overline{y} \\
\Delta(q_2,a) & \equiv & q_2 \wedge x > \overline{x} \wedge y > \overline{y} \\
\Delta(q_2,b) & \equiv & q_4 \wedge \overline{x}>\overline{y} \\\\
\end{array}\]
\centerline{(b)}
\end{minipage}
\end{center}
\caption{Alternating Data Automaton Example}
\label{fig:ex}
\end{figure}

@@

In this paper we tackle the following problems: \begin{compactenum}
\item \emph{Boolean closure}: given automata $\A_1$ and $\A_2$, both
  with the same set of variables $\vec{x}$, do there exist automata
  $\A_\cup$, $\A_\cap$ and $\overline{\A_1}$ such that $L(\A_\cup) =
  \A_1 \cup \A_2$, $L(A_\cap) = \A_1 \cap \A_2$ and
  $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$ ?
\end{compactenum}

@@

Given a set $Q$ of Boolean variables and a set $\vec{x}$ of variables
of sort $\Data$, for a formula $\phi \in \Form^+(Q,\vec{x})$, with no
negated occurrences of the Boolean variables, we define the formula
$\overline{\phi} \in \Form^+(Q,\vec{x})$ recursively on the structure
of $\phi$:
\[\begin{array}{rclcrcl}
\overline{\phi_1 \vee \phi_2} & \equiv & \overline{\phi_1} \wedge \overline{\phi_2} && 
\overline{\phi_1 \wedge \phi_2} & \equiv & \overline{\phi_1} \vee \overline{\phi_2} \\ 
\overline{\neg\phi} & \equiv & \neg \overline{\phi} \text{ if $\phi$ not atom} &&
\overline{\phi} & \equiv & \phi \text{ if $\phi \in Q$} \\
\overline{\phi} & \equiv & \neg\phi \text{ if $\phi \not\in Q$ atom}
\end{array}\]
We have $\len{\overline{\phi}} = \len{\phi}$, for every formula $\phi
\in \Form^+(Q,\vec{x})$.

@@

It is easy to see that $\len{\A_\cup} = \len{\A_\cap} = \len{\A_1} +
\len{\A_2}$ and $\len{\overline{\A}} = \len{\A}$, thus the automata
for the Boolean operations, including complementation, can be built in
linear time. This matches the linear-time bounds for intersection and
complementation of alternating automata over finite alphabets
\cite{ChandraKozenStockmeyer81}.

@@

Since no Boolean variable from $Q$ occurs under negation in $\phi$, it
is easy to prove the following monotonicity property: given two
formulae $\phi,\psi \in \Form^+(Q,\vec{x})$ if $\phi \models \psi$
then $\Post{\A}(\phi,u) \models \Post{\A}(\psi,u)$, for any $u \in
\Sigma^*$. This suggest an improvement of the above semi-algorithm,
that enumerates and stores only a set $U \subseteq \Sigma^*$ for which
$\set{\Post{\A}(\phi,u) \mid u \in U}$ forms an
\emph{antichain}\footnote{Given a partial order $(D,\preceq)$ an
  antichain is a set $A \subseteq D$ such that $a \not\preceq b$ for
  any $a,b \in A$.} w.r.t. the entailment partial order. This is
because, for any $u,v \in \Sigma^*$, if $\Post{\A}(\iota,u) \models
\Post{\A}(\iota,v)$ and $\Accept{\A}(uw)$ is satisfiable for some $w
\in \Sigma^*$, then $\Post{\A}(\iota,uw) \models \Post{\A}(\iota,vw)$,
thus $\Accept{\A}(vw)$ is satisfiable as well, and there is no need
for $u$, since the non-emptiness of $\A$ can be proved using $v$
alone. However, even with this optimization, the enumeration of
sequences from $\Sigma^*$ diverges in many real cases, because
infinite antichains exist in many interpretations, e.g.\ $q \wedge x
\teq 0,~ q \wedge x \teq 1, \ldots$ for $\Data^\I = \nat$.

A \emph{safety invariant} for $\A$ is a function $\Inv : (Q \mapsto
\booli) \rightarrow 2^{\vec{x} \mapsto \Data^\I}$ such that, for every
Boolean valuation $\beta : Q \rightarrow \booli$, every valuation $\nu
: \vec{x} \mapsto \Data^\I$ of the data variables and every finite
sequence $u \in \Sigma^*$ of input events, the following
hold: \begin{compactenum}
%% \item $\beta \models \iota \Rightarrow \nu_0 \in \Inv(\beta)$, where
%%   $\nu_0(x) = \mathbf{0}$, for all $x \in \vec{x}$.
%
\item\label{it1:inv} $\I,\beta \cup \nu \models \Post{\A}(\iota,u)
  \Rightarrow \nu \in \Inv(\beta)$, and
%
\item\label{it2:inv} $\nu \in \Inv(\beta) \Rightarrow \I,\beta \cup
  \nu \not\models \Accept{\A}(u)$.
\end{compactenum}
If $\Inv$ satisfies only the first point above, we call it an
\emph{invariant}. Intuitively, a safety invariant maps every Boolean
valuation into a set of data valuations, that contains the initial
configuration $\iota \equiv \Post{\A}(\iota,\varepsilon)$, whose data
variables are unconstrained, over-approximates the set of reachable
valuations (point \ref{it1:inv}) and excludes the valuations satisfying the
acceptance condition (point \ref{it2:inv}).  A formula $\phi(Q,\vec{x})$
is said to \emph{define} $\Inv$ iff for all $\beta: Q \rightarrow \booli$ and
$\nu : \vec{x} \rightarrow \Data^\I$, we have $\I,\beta\cup\nu \models
\phi$ iff $\nu \in \Inv(\beta)$.

@@

We are left with fulfilling point (\ref{it2:inv}) from the definition
of a safety invariant. To this end, suppose that, for a given set
$\preds$ of predicates, the invariant $\abs{\Inv}$, defined by the
previous lemma, meets point (\ref{it1:inv}) but not point
(\ref{it2:inv}), where $\Post{\A}$ and $\Accept{\A}$ replace
$\AbsPost{\A}$ and $\AbsAccept{\A}$, respectively. In other words,
there exists a finite sequence $u \in \Sigma^*$ such that $\nu \in
\abs{\Inv}(\beta)$ and $\I,\beta\cup\nu \models \AbsAccept{\A}(u)$,
for some Boolean $\beta : Q \rightarrow \booli$ and data $\nu :
\vec{x} \rightarrow \Data^\I$ valuations. Such a $u\in\Sigma^*$ is
called a \emph{counterexample}.

@@

Given an unsatisfiable conjunction of formulae $\psi_1 \wedge \ldots
\wedge \psi_n$, an \emph{interpolant} is a tuple of formulae
$\tuple{I_1,\ldots,I_{n-1},I_n}$ such that $I_n \equiv \bot$, $I_i
\wedge \psi_i \models_\T I_{i+1}$ and $I_i$ contains only variables
and function symbols that are common to $\psi_i$ and $\psi_{i+1}$, for
all $i \in [n-1]$. Moreover, by Lyndon's Interpolation Theorem
\cite{Lyndon59}, we can assume without loss of generality that every
Boolean variable with at least one positive (negative) occurrence in
$I_i$ has at least one positive (negative) occurrence in both $\psi_i$
and $\psi_{i+1}$. In the following, we shall assume the existence of
an interpolating decision procedure for $\theory(\sig,\I)$ that meets
the requirements of Lyndon's Interpolation Theorem.

@@

Given a non-empty spurious counterexample $u = a_1\ldots a_n$, where
$n>0$, we consider the following interpolation problem: 
\begin{eqnarray}\label{eq:interpolation-problem}
\Theta(u) & \equiv & \theta_0(Q_0) \wedge \theta_1(Q_0 \cup Q_1,\vec{x}_0
\cup \vec{x}_1) \wedge \ldots \\ 
&& \wedge~ \theta_n(Q_{n-1} \cup Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge \theta_{n+1}(Q_n) \nonumber
\end{eqnarray}
where $Q_k = \set{q_k \mid q \in Q}$, $k \in [0,n]$ are time-stamped
sets of Boolean variables corresponding to the set $Q$ of states of
$\A$. The first conjunct $\theta_0(Q_0) \equiv \iota[Q_0/Q]$ is the
initial configuration of $\A$, with every $q \in \fv{\Bool}{\iota}$
replaced by $q_0$. The definition of $\theta_k$, for all $k\in[1,n]$,
uses \emph{replacement sets} $R_\ell \subseteq Q_\ell$, $\ell\in
[0,n]$, which are defined inductively below: \begin{compactitem}
\item $R_0 = \fv{\Bool}{\theta_0}$,
%
\item $\theta_\ell \equiv \bigwedge_{q_{\ell-1}\in R_{\ell-1}}
  (q_{\ell-1} \rightarrow
  \Delta(q,a_\ell)[Q_\ell/Q,\vec{x}_{\ell-1}/\overline{\vec{x}},\vec{x}_\ell/\vec{x}])$
  and $R_\ell = \fv{\Bool}{\theta_\ell} \cap Q_\ell$, for each
  $\ell\in[1,n]$.
%
\item $\theta_{n+1}(Q_n) \equiv \bigwedge_{q \in Q \setminus F} (q_n
  \rightarrow \bot)$.
\end{compactitem}
The intuition is that $R_0,\ldots,R_n$ are the sets of states
replaced, $\theta_0, \ldots, \theta_n$ are the sets of transition
rules fired on the run of $\A$ over $u$ and $\theta_{n+1}$ is the
acceptance condition, which forces the last remaining non-final states
to be false.
%
%% Moreover, we require that an interpolant
%% $\tuple{\top,I_0,\ldots,I_n,\bot}$ for the interpolation problem
%% $\Theta(u)$ does not have negative occurrences of states, i.e.\ $I_i
%% \in \Form^+(Q,\vec{x})$, for all $i \in [0,n]$. Such an interpolant
%% can always be built, as showed below:
%
%% \begin{proposition}\label{prop:positive-interpol}
%%   If $\Theta(a_1 \ldots a_n)$ is unsatisfiable then one can build an
%%   interpolant $\tuple{\top,I_0,\ldots,I_n,\bot}$ for $\Theta(a_1\ldots
%%   a_n)$, such that $I_i \in \Form^+(Q,\vec{x})$, for all $i \in
%%   [0,n]$.
%% \end{proposition}
%
We recall that a run of $\A$ over $u$ is a sequence: \[\phi_0(Q)
\Rightarrow \phi_1(Q,\vec{x}_0\cup\vec{x}_1) \Rightarrow \ldots
\Rightarrow \phi_n(Q,\vec{x}_0\cup\ldots\cup\vec{x}_n)\] where
$\phi_0$ is the initial configuration $\iota$ and for each $k>0$,
$\phi_k$ is obtained from $\phi_{k-1}$ by replacing each state $q \in
\fv{\Bool}{\phi_{k-1}}$ by the formula
$\Delta(q,a_k)[\vec{x}_{k-1}/\overline{\vec{x}},\vec{x}_k/\vec{x}]$,
given by the transition function of $\A$. Observe that, because the
states are replaced with transition formulae when moving one step in a
run, these formulae lose track of the control history and are not
suitable for producing interpolants that relate states and data.

The main idea behind the above definition of the interpolation problem
is that we would like to obtain an interpolant $\tuple{\top,I_0(Q),
  I_1(Q,\vec{x}), \ldots, I_{n}(Q,\vec{x}),\bot}$ whose formulae
\emph{combine states with the data constraints that must hold
  locally}, whenever the control reaches a certain Boolean
configuration. This association of states with data valuations is
tantamount to defining efficient semi-algorithms, based on lazy
abstraction \cite{HJMS02}. Furthermore, the abstraction defined by the
interpolants generated in this way can also \emph{over-approximate the
  control structure} of an automaton, in addition to the sets of data
values encountered throughout its runs.

@@

Termination of Algorithm \ref{alg:predabs} depends on the ability of a
given interpolating decision procedure for the combined Boolean and
data theory $\theory(\sig,\I)$ to provide interpolants that yield a
safety invariant, whenever $L(\A) = \emptyset$. In this case, we use
the covering relation $\lhd$ to ensure that, when a newly generated
node is covered by a node already in $N$, it is not added to the
worklist, thus cutting the current branch of the search.

@@

In this process, the nodes on the path between $\rootNode$ and $n$,
including $n$, might become eligible for coverage, therefore we
attempt to close each ancestor of $n$ that is impacted by the
refinement (line \ref{ln:refine-end}). Observe that, in this case the
call to $\Call{Close}{}$ must uncover each node which is covered by a
successor of $n$ (line \ref{ln:close-uncover} of the $\Call{Close}{}$
function). This is required because, due to the over-approximation of
the sets of reachable configurations, the covering relation is not
transitive, as explained in \cite{mcmillan06}. If $\Call{Close}{}$
adds a covering edge $(n_i,m)$ to $\lhd$, it does not have to be
called for the successors of $n_i$ on this path, which is handled via
the Boolean flag $b$.
